---
title: "Homework 2 - `ggplot2`"
author: "your_name_here"
date: "2023-09-08"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.width = 6, 
                      fig.height = 6, 
                      fig.align = 'center')
```

## Question 1

The data frame `diamonds`, like `mpg`, is included in the `ggplot2` package. This data records the attributes of several thousand diamonds sold by a wholesale online retailer. For this question, your goal is to recreate the graph shown below as closely as possible. A few hints:

  + Pay attention to scales, theme, and labels
  + Find what transformations are available for continuous x axis
  + The argument `alpha = 0.3` is used in one of the layers to give each point 30% opacity
  + Default colors are used

```{r}
library(ggplot2)
data("diamonds")
```

```{r, echo = FALSE, cache=TRUE}
ggplot(diamonds, aes(carat,  price, color = color)) +
  geom_point(alpha = 0.3) + theme_bw() + scale_x_continuous(trans = "log2") +
  labs(x = "Carat", y = "Sale Price ($)", color = "Color Grade", title = "Diamond Sales") + theme(legend.position = "bottom")
```


## Question 2

## Question 3

## Question 4
